Learning Discrete Bayesian Networks from Continuous Data

### 2017-05-10 jair_Learning_Discrete_Bayesian_Networks_from_Continuous_Data.tex
compile fixes

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\usepackage[round]{natbib}
\usepackage{multirow}
\usepackage{pgfplots}
\usepackage{pgfplotstable}
\usepackage{tikz}
\usetikzlibrary{arrows.meta}
\usepgfplotslibrary{groupplots}
\usepackage{lscape}
\usepackage{siunitx}
\usepackage{booktabs}
\usepackage{BOONDOX-cal} % for lowercase caligraphic
\usepackage[font=small,skip=5pt]{caption}
\usepackage[hyphens]{url}
\usepackage{subcaption}
\captionsetup{compatibility=false}
\usepackage{tabularx}

@@

\usetikzlibrary{bayesnet}
\setcitestyle{aysep={ }} %Author-Year Separator
%
% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
%
% insert here the call for the packages your document requires
%\usepackage{latexsym}
% etc.
%
% please place your own definitions here and don't use \def but
% \newcommand{}{}
%
% Insert the name of "your journal" with
% \journalname{myjournal}
%

@@

Bayesian networks \citep{Pearl_1988, PGM_2009} are often used to model uncertainty and causality, with applications ranging from decision-making systems \citep{kochenderfer2012next} to medical diagnosis \citep{Lustgarten_2011}. Learning Bayesian networks from raw data can help users to discover knowledge without field-specific experts, such as causality and probabilistic relation between variables. It is common to assume that the random variables in a Bayesian network are discrete, since many Bayesian network learning algorithms are unable to efficiently handle continuous variables. In addition, many of the commonly used Bayesian network software packages, such as Netica \citep{netica1992}, SMILearn \citep{druzdzel1999smile}, and bnlearn \citep{bnlearn2010}, are geared towards discrete variables. However, many applications require the use of continuous data, such as position and velocity in dynamic systems \citep{kochenderfer2010airspace}.

@@

The third approach is discretization.
Automated discretization methods have been studied in machine learning and statistics for many years \citep{Dougherty_1995, Kerber_1992, Holte_1993, Fayyad_1993}, primarily for classification problems.
These methods learn the best discretization policy for a continuous attribute by considering its interaction with a class variable.
It is common to discretize all continuous variables before learning a Bayesian network structure to avoid having to consider variable interactions, as the interactions and dependencies between variables in Bayesian networks introduce complexity.
Prior work exists for discretizing continuous variables in naive Bayesian networks and tree-augmented networks \citep{Fried_naive}, but only a few discretization methods for general Bayesian networks have been proposed \citep{Friedman_1996, Kozlov_1997, Monti_1998, Steck_2007}.

A common discretization method for Bayesian networks is to split continuous data into uniform-width intervals or to use field-specific expertise.
Several Bayesian methods exist to discretize according to the marginal distribution \citep{scott1979optimal, freedman1981histogram, knuth2013optimal, scargle2013studies}, but this ignores dependencies between variables and leads to suboptimal performance, as demonstrated in Appendix A.
A more principled method is the minimum description length (MDL) principle discretization \citep{Friedman_1996}.
The MDL principle proposed by \cite{MDL_1978} states that the best model for a dataset is the one that minimizes the amount of information needed to describe it \citep{Grunwald_2009}.
MDL methods trade off goodness-of-fit against model complexity to reduce generalization error.
In the context of Bayesian networks, \cite{Friedman_1996} applied the MDL principle to determine the optimal number of discretization intervals for continuous variables and the optimal positions of their discretization edges.
Their approach selects a discretization policy that minimizes the sum of the description length of the discretized Bayesian network and the information necessary for recovering the continuous values from the discretized data.

@@

It is often necessary to infer the structure of a Bayesian network from data.
Three common approaches to Bayesian network structure learning are constraint-based, score-based, and Bayesian model averaging \citep[see][chap.~18]{PGM_2009}.
This work uses the K2 structure learning algorithm \citep{cooper1992bayesian}, a frequently used score-based structure learning method.
Score-based structure learning methods over discrete variables commonly evaluate candidate structures according to their likelihood against a dataset $D$.
In practice, one maximizes the log-likelihood, also known as the Bayesian score, $\ln P(G\mid D)$~\citep{cooper1992bayesian}.

@@

This section illustrates the advantage of structure-aware discretization methods over marginal discretization methods \citep{scott1979optimal, freedman1981histogram, knuth2013optimal, scargle2013studies}.
Marginal discretization methods ignore intervariable dependencies and can thus lead to suboptimal discretizations for the joint distribution.
Consider a synthetic dataset with two variables, discrete $A$ and continuous $B$, with \num{500} data samples sampled from the Bayesian network $A \rightarrow B$ described in Table~8.  %~\ref{table:marginal_counter}.
Following the experiment procedure in Section~\ref{sec:experiments}, both the MDL method and the proposed Bayesian approach find the true discretization edge for $B$ near $1.0$ and have log likelihood \num{-0.023}.
They recover the discretization shown in Figure~\ref{fig:marginal_counter}.
The Bayesian Blocks method~\citep{scargle2013studies}, as with any marginal method, cannot distinguish between the different conditional distributions and thus does not locate a discretization edge.
The resulting log likelihood is worse, at \num{-0.69}.
This example shows that structure-aware discretization methods can leverage additional information and thus produce better joint representations.

>>>>

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\usepackage{multirow}
\usepackage{pgfplots}
\usepackage{pgfplotstable}
\usepackage{tikz}
\usetikzlibrary{arrows.meta}
\usepgfplotslibrary{groupplots}
\usepackage{lscape}
\usepackage{siunitx}
\usepackage{booktabs}
\usepackage{BOONDOX-cal} % for lowercase caligraphic
\usepackage[font=small,skip=5pt]{caption}
\usepackage[hyphens]{url}
\usepackage{subcaption}
\captionsetup{compatibility=false}
\usepackage{tabularx}

@@

\usetikzlibrary{bayesnet}

\newcommand{\citep}[1]{\cite{#1}}
\newcommand{\citet}[1]{\citeauthor{#1} \citeyear{#1}}

%
% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
%
% insert here the call for the packages your document requires
%\usepackage{latexsym}
% etc.
%
% please place your own definitions here and don't use \def but
% \newcommand{}{}
%
% Insert the name of "your journal" with
% \journalname{myjournal}
%

@@

Bayesian networks \citep{Pearl_1988,PGM_2009} are often used to model uncertainty and causality, with applications ranging from decision-making systems \citep{kochenderfer2012next} to medical diagnosis \citep{Lustgarten_2011}. Learning Bayesian networks from raw data can help users to discover knowledge without field-specific experts, such as causality and probabilistic relation between variables. It is common to assume that the random variables in a Bayesian network are discrete, since many Bayesian network learning algorithms are unable to efficiently handle continuous variables. In addition, many of the commonly used Bayesian network software packages, such as Netica \citep{netica1992}, SMILearn \citep{druzdzel1999smile}, and bnlearn \citep{bnlearn2010}, are geared towards discrete variables. However, many applications require the use of continuous data, such as position and velocity in dynamic systems \citep{kochenderfer2010airspace}.

@@

The third approach is discretization.
Automated discretization methods have been studied in machine learning and statistics for many years \citep{Dougherty_1995,Kerber_1992,Holte_1993,Fayyad_1993}, primarily for classification problems.
These methods learn the best discretization policy for a continuous attribute by considering its interaction with a class variable.
It is common to discretize all continuous variables before learning a Bayesian network structure to avoid having to consider variable interactions, as the interactions and dependencies between variables in Bayesian networks introduce complexity.
Prior work exists for discretizing continuous variables in naive Bayesian networks and tree-augmented networks \citep{Fried_naive}, but only a few discretization methods for general Bayesian networks have been proposed \citep{Friedman_1996,Kozlov_1997,Monti_1998,Steck_2007}.

A common discretization method for Bayesian networks is to split continuous data into uniform-width intervals or to use field-specific expertise.
Several Bayesian methods exist to discretize according to the marginal distribution \citep{scott1979optimal,freedman1981histogram,knuth2013optimal,scargle2013studies}, but this ignores dependencies between variables and leads to suboptimal performance, as demonstrated in Appendix A.
A more principled method is the minimum description length (MDL) principle discretization \citep{Friedman_1996}.
The MDL principle proposed by \cite{MDL_1978} states that the best model for a dataset is the one that minimizes the amount of information needed to describe it \citep{Grunwald_2009}.
MDL methods trade off goodness-of-fit against model complexity to reduce generalization error.
In the context of Bayesian networks, \cite{Friedman_1996} applied the MDL principle to determine the optimal number of discretization intervals for continuous variables and the optimal positions of their discretization edges.
Their approach selects a discretization policy that minimizes the sum of the description length of the discretized Bayesian network and the information necessary for recovering the continuous values from the discretized data.

@@

It is often necessary to infer the structure of a Bayesian network from data.
Three common approaches to Bayesian network structure learning are constraint-based, score-based, and Bayesian model averaging \citep{PGM_2009}.
This work uses the K2 structure learning algorithm \citep{cooper1992bayesian}, a frequently used score-based structure learning method.
Score-based structure learning methods over discrete variables commonly evaluate candidate structures according to their likelihood against a dataset $D$.
In practice, one maximizes the log-likelihood, also known as the Bayesian score, $\ln P(G\mid D)$~\citep{cooper1992bayesian}.

@@

This section illustrates the advantage of structure-aware discretization methods over marginal discretization methods \citep{scott1979optimal,freedman1981histogram,knuth2013optimal,scargle2013studies}.
Marginal discretization methods ignore intervariable dependencies and can thus lead to suboptimal discretizations for the joint distribution.
Consider a synthetic dataset with two variables, discrete $A$ and continuous $B$, with \num{500} data samples sampled from the Bayesian network $A \rightarrow B$ described in Table~8.  %~\ref{table:marginal_counter}.
Following the experiment procedure in Section~\ref{sec:experiments}, both the MDL method and the proposed Bayesian approach find the true discretization edge for $B$ near $1.0$ and have log likelihood \num{-0.023}.
They recover the discretization shown in Figure~\ref{fig:marginal_counter}.
The Bayesian Blocks method~\citep{scargle2013studies}, as with any marginal method, cannot distinguish between the different conditional distributions and thus does not locate a discretization edge.
The resulting log likelihood is worse, at \num{-0.69}.
This example shows that structure-aware discretization methods can leverage additional information and thus produce better joint representations.
